### 2023-01-27 paper-thesis/chapters/21-adding-inference.tex
Improve inference chapter

A more powerful algorithm that could solve this is \emph{higher-order pattern unification} as defined by Miller~\cite{Miller89}, which is implemented in Agda. Implementing this in Statix is theoretically possible\footnote{since Statix is turing-complete}, but would be quite difficult to do in practice.

@@

\section{Implementing AFOU}
\label{implementing-inference}

@@

	\item A complex expression \verb|e1| on the left, an infer expression on the right. In this case, we know what top-level constructor of the metavariable should be, but not necessarily the entire constructor (there might be infers in \verb|e1|). We introduce new \verb|Infer| expressions on the left, and call \verb|expectBetaEq| recursively. A simple example for \verb|BoolIf| is below:
	\begin{lstlisting}
expectBetaEq((s1, e1@BoolIf(c1, t1, b1)), (_, Infer(e2))) :-
	{c2 t2 b2}
	e2 == BoolIf(Infer(c2), Infer(t2), Infer(b2)),
	expectBetaEq((s1, e1), (sEmpty(), e2)).
	\end{lstlisting}
	\verb|FnType| and \verb|FnConstruct| work similarly,

@@

	\item \label{inf-case4} An infer expression on both sides. This is where the approximation comes in. In normal first-order unification, we would see if either side is known, and possible apply one of the rules above depending on the result. This is not possible, so we're going to do something that approximates first-order unification: just set both sides to be equal. This is an approximation because this might fail if both sides are equal under beta equality but not identical. This approximation is analyzed in section \ref{analysing-inference}.
\begin{lstlisting}
expectBetaEq((_, Infer(e1)), (_, Infer(e2))) :-
	e1 == e2.
\end{lstlisting}

@@

The only difference between AFOU and FOU is case \ref{inf-case4}. Instead of asserting that both sides are equal under beta equality, we assert that both sides are identical. Both sides being identical implies that they are beta-equal, so this approximation is sound, meaning there is no program that AFOU can infer but FOU cannot.

The approximation is not complete: There are situations where AFOU fails to infer a program that FOU can infer. However, these programs are rare. This happens when a program contains multiple infer points, two of these are asserted to be beta-equal, and they are not identical but they are beta-equal. \todo{Find program where this fails.}

>>>>

A more powerful algorithm that could solve this is \emph{higher-order pattern unification} as defined by Miller~\cite{Miller89}, which is implemented in Agda. Implementing this in Statix is theoretically possible\footnote{Since Statix is turing-complete}, but would be quite difficult to do in practice.

@@

\section{Implementing AFOU} \todo{Do we want all these examples in Statix or math notation?}
\label{implementing-inference}

@@

	\item A complex expression \verb|e1| on the left, an infer expression on the right. In this case, we know what top-level constructor of the metavariable should be, but not necessarily the entire constructor (there might be infers in \verb|e1|). We introduce new \verb|Infer| expressions on the left, and call \verb|expectBetaEq| recursively. A simple example for \verb|BoolIf| is below:
	\begin{lstlisting}
expectBetaEq((s1, e1@BoolIf(c1, t1, b1)), (_, Infer(e2))) :-
	e2 == BoolIf(Infer(c2), Infer(t2), Infer(b2)),
	expectBetaEq((s1, e1), (sEmpty(), e2)).
	\end{lstlisting}
	\verb|FnType| and \verb|FnConstruct| work similarly, but we don't only have sub-expressions but also a name to consider. Sadly, this is the first case where we have to approximate. 
	\begin{lstlisting}
expectBetaEq((s1, e1@FnConstruct(arg_name1, arg_type1, body1)), 
	(_, Infer(e2))) :-
	e2 == FnConstruct(arg_name1, Infer(arg_type2), Infer(body2)),
	expectBetaEq_((s1, e1), (sEmpty(), e2)).
	\end{lstlisting}
	\label{inf-case2}
	Ideally, we would generate a new name iff \verb|e2| is a free variable, otherwise using the already generated name. We can't query whether \verb|e2| is free, so as a best-attempt we always assume that \verb|e2| is free and generate a new name, this will fail in some cases. We will discuss this in the next section.

@@

	\item \label{inf-case4} An infer expression on both sides. Here, more approximation is required. In normal first-order unification, we would see if either side is known, and possible apply one of the rules above depending on the result. This is not possible, so we're going to do something that approximates first-order unification: just set both sides to be equal. This is an approximation because this might fail if both sides are equal under beta equality but not identical. This approximation is analyzed in section \ref{analysing-inference}.
\begin{lstlisting}
expectBetaEq((_, Infer(e1)), (_, Infer(e2))) :-
	e1 == e2.
\end{lstlisting}

@@

The only difference between AFOU and FOU is case \ref{inf-case2} (for functions) and \ref{inf-case4}.

In case \ref{inf-case2}, this will only infer the value if it has not been inferred yet. This means that if the same value is inferred from two different places, this will infer two different values and fail.

In case \ref{inf-case4}, instead of asserting that both sides are equal under beta equality, we assert that both sides are identical. Both sides being identical implies that they are beta-equal, so this approximation is sound, meaning there is no program that AFOU can infer but FOU cannot.

The approximations are not complete: There are situations where AFOU fails to infer a program that FOU can infer. However, these programs are surprisingly uncommon considering how rough the approximations are. They mostly fail when we attempt to infer function values, and two functions are inferred that are beta-equal, and they are beta-equal but not identical. An example of a program where these approximation fails is:
